Smart Representations: {R}ationality, Regret, Evolution \& the Environment

### 2015-11-06 paper/paper.tex
tweaks and extracted bib
Paolo fixes Sec 3.3

\usepackage[]{helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
% \bibliography{helpers/MyRefLocal}
\bibliography{MyRefGlobal}
\usepackage{helpers/myenvironments}
\usepackage{helpers/mycommands}

@@

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a set of
players, $A_i$ a set of acts and $\pi_i$ a payoff function for each player $i \in N$. For the
most part, this paper will focus on symmetric games with two players. A two-player game is
symmetric if players have the same action set, $A_1 = A_2$, and payoffs are symmetric:
$\pi_1(\myact{k}, \myact{l}) = \pi_2(\myact{l}, \myact{k})$ for all $k$ and $j$. An example is
given in Figure~\ref{coordgame1}, which shows a coordination game. The numbers give the payoff
for the player choosing a row act against an opponent choosing a column act. Since the game is
symmetric, it suffices to give just these numbers.

@@

More generally even, a choice mechanism could be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to keep matters simple and implementable for a start, we will here only look at choice
mechanisms that differ only with respect to their subjective representation function $\gamma$,
which in turn is conceived of as a \emph{uniform} way of forming preference and belief
\emph{types}. \textcolor{red}{Formally, $\gamma$ is a function in
  $\mathcal{G} \rightarrow \Theta \times E$, where \dots. Then, given a set of subjective
  preferences $ \Theta $ [@Paolo: minimally define subjective preferences somewhere above] and
  a set of epistemic states (or belief types) $E$, a choice mechanism is a function
  $\kappa: \Theta \times E \rightarrow A_i $.}

\textcolor{red}{More concretely, we focus on two \emph{preference types} and two \emph{belief
    types}. We view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and
  a belief type as a general disposition to form beliefs about the co-player’s behavior. The
  two possible dispositions that we take account of are probabilistic (precise) belief and
  non-probabilistic (imprecise) belief, as we will explain in the next section.}

\textcolor{red}{We are most interested here in studying whether evolutionary pressure could
  support non-veridical subjective preferences under certain conditions. We therefore consider
  two \emph{preference types}: veridical type $\tau^\pi$, who uses the actual payoff function
  $\pi$ of the actual game $G$, and regret type $\tau^{reg}$, who looks at the regret
  transformation of the actual game $G$.}

\textcolor{red}{The choice mechanism that we use in this paper is \textit{maxmin expected
    utility} [GS89], for reasons that will be clear and justified in the following two
  sections.  }

@@

In evolutionary game theory, for instance, players in a population obviously face uncertainty about the composition of the population that they are part of, and consequently about the co-player that they are randomly paired with at each round and about the co-player's action. We represent this uncertainty in two different ways that correspond to the two approaches discussed above: a Bayesian uniform distribution and an imprecise probability set. This amounts to having two different belief types, $ \overline{\mu}$ and $ \Delta(A_{-i}) \rbrace $, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

@@

In decision theory, the imprecise probability model is in line with most of the representation results in decision making under uncertainty [GilboaSchmeidler89,KMM06], and seems justified by empirical observations too.

@@

The pattern consistently observed in experimets is: $f_{{\color{red}R}} \succ f_{{\color{yellow}Y}}$ and $f_{{\color{red}R}B} \prec f_{{\color{yellow}Y}B}$. This is clearly incompatible with any precise probabilistic belief about the proportion of black and yellow marbles. This kind of behavior has been
famously axiomatized by Gilboa and Schmeidler in a paper from 1989. The choice pattern is normally explained in terms of aversion to uncertainty, and the standard representation is given by means of probability sets together with maxmin rule. We can think of the DM in Ellsberg's example as maximinimizing expected utility over the probability set $ [\underline{P}(R)=\frac{1}{3},\underline{P}(Y)=0,\underline{P}(B)=0,\overline{P}(R)=\frac{1}{3},\overline{P}(Y)=\frac{2}{3},\overline{P}(B)=\frac{2}{3}]  $. This amounts to calculating for any probability distribution contained in the probability set the expected utility of any available action, and then to choosing the action that guarantees the highest minimal expected utility. The following observation is then trivial.

@@

Evidence from experimental literature suggests that agents are mainly uncertainty
averse [REF]. In line with empirical data,
we assume that players in the population are uncertainty averse and conform to maxmin expected
utility. The resulting behavior is then produced by maximinimizing the subjective utility
given by the player (preference, UNIFORM TERMINOLOGY CHECK) type over the set of probabilities given by the belief type of the agent.

@@

Fixing the choice mechanism (action selection component) to maximin expected utility, and allowing two belief and
two preference types, we regain all the choice rules considered in sectionNN from a more general and unified standpoint, as it is summed up in
Table~\ref{tab:CMs}. \\
We are now in the position of formally deriving behavior from a pair $(\theta, e) \in \Theta \times E$.

@@

Given a game $G$ and a player $(\tau, e)$, we define $\text{maxmin}(\tau(G), e):= \lbrace a'_i \in A_i : a_i \in \text{max}_{a_i \in A_i} \text{ min}_{\mu \in e} \text{ } \tau(G)(a_i, \mu) \rbrace$.

@@

[[\noindent Definition \ref{def:expected fitness} gives the expected evolutionary fitness of player $(\tau, e)$ against player $(\tau', e')$ in game $G$.\footnote{As already specified in the previous example, we assume that, whenever the set $\text{maxmin}(\tau(G), e)$ is not a singleton, the player $(\tau, e)$ will mix with equal probability among the actions in $\text{maxmin}(\tau(G), e)$.}
[OF COURSE we will postpone the definition of evolutionary fitness, it is only momentarily here!]
\begin{definition} \label{def:expected fitness}

Given a game $G \in \mathcal{G}$, a player $(\tau, e)$, and a player $(\tau', e')$, we define $\pi^G((\tau, e),(\tau', e')):= \pi^G(\text{maxmin}(\tau(G), e),\text{maxmin}(\tau'(G), e'))$.

@@

It is then worth stressing here that $\kappa(\tau^{\pi}, \Delta(A_{-i}))$ turns out to be the objective
maximinimizer, $\kappa(\tau^{reg}, \Delta(A_{-i}))$ corresponds to the regret minimizer,
$\kappa(\tau^{\pi}, \overline{\mu})$ is the objective expected utility maximizer (with flat beliefs), and
$\kappa(\tau^{reg}, \overline{\mu})$ is the expected regret minimizer (with flat beliefs).
Moreover, different subjective preferences and belief types can give rise to identical action predictions,
either in a restricted class of games, or in general. This is also the case here.

@@

\noindent In other words, given the same probabilistic belief, maximization of expected
(objective) payoff and minimization of expected (subjective) regret are behaviorally
equivalent. This holds for arbitrary games, as well as for solitary decision situations. On the
other hand, the following does not hold in general:

@@

Many other alternatives different from maxmin could produce behavior: choice rules like minmax, maxmax, and minmin would similarly generate behavior in case of a lack of precise probabilistic belief. Here we postulate that players use maxmin for two main reasons that we will expand in the next section. First, among other possible choice rules maxmin is not only intuitively appealing but also corroborated from an empirical point of view. Secondly, as it will be clear in the following, our main interest is the evolution of different patterns of subjective conceptualizations of games (player types), and we want to show that the veridical representation (objective type) is not necessarily the fittest and may not be evolutionarily stable. Assuming that all players use a fixed rule (maxmin) independent of their player type allows us to factor the choice rule out of the selection process and to focus on player types.

@@

\todo[inline]{PAOLO rewrite up to here!!! BUT the last 2-3 paragraphs must be rearranged a bit more!}

@@

In sum, we have so far considered four different alternatives, each differing along a
preference and a belief dimension. As for preference representations, agents could either
target objective payoffs, or consider non-veridical subjective regret transformations. Given a game $G$, we will denote by $ \tau^{reg}(G)$ the regret representation (corresponding to the negative regret transformation) of the game, and $\tau^{\pi}(G)$ the objective representation (corresponding to the identity transformation) of the game. \\
As for
beliefs, agents could either hold probabilistic belief or not. In the former case, we would
assume that they maximize expectations under whatever preference representation they hold; in
the latter, we would assume that they play maximin based on their preference
representations. There are doubtlessly many more possible preference and belief types, but
these are the main characters in the plot to follow.

Before we turn to evolutionary competition between choice principles, we should justify our
focus on the two choice rules. In the following we argue that both the minimax rule, as well as the maximization of expected utility 
(and the Laplace rule as a special case) can be viewed as the same choice rule under different representations of
uncertainty.

To sum up, here we consider the choice of an action as depending on two features: the subjective preference and the belief state. Definition \ref{def:choice rule} makes this point clear.

\begin{definition} \label{def:choice rule}

Given a set of subjective preferences $\Theta$ and a set of epistemic types $E$, we define \textit{choice rule} a function $\kappa: \Theta \times E  \rightarrow A_i$ that associates an action choice to a pair $(\theta, e) \in \Theta \times E$.

According to what we argued in this section, we will assume that the choice rule that turns subjective preferences and beliefs into actions is maxmin expected utility.

>>>>

\usepackage[]{helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
% \bibliography{helpers/MyRefLocal}
\bibliography{choicePrince}
\usepackage{helpers/myenvironments}
\usepackage{helpers/mycommands}

@@

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a finite set of
players, $A_i$ a finite set of acts and $\pi_i$ a payoff function for each player $i \in N$. For the
most part, this paper will focus on symmetric games with two players. A two-player game is
symmetric if players have the same action set, $A_1 = A_2$, and payoffs are symmetric:
$\pi_1(\myact{k}, \myact{l}) = \pi_2(\myact{l}, \myact{k})$ for all $k$ and $j$. An example is
given in Figure~\ref{coordgame1}, which shows a coordination game. The numbers give the payoff
for the player choosing a row act against an opponent choosing a column act. Since the game is
symmetric, it suffices to give just these numbers.

@@

More generally even, a choice mechanism could be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to keep matters simple and implementable for a start, we will here only look at choice
mechanisms that differ only with respect to their subjective representation function $\gamma$,
which in turn is conceived of as a \emph{uniform} way of forming preference and belief
\emph{types}.

Formally, given a class $ \mathcal{G} $ of games, $\gamma: \mathcal{G} \rightarrow \Theta \times E$ is a function that assigns a subjective preference $\theta \in \Theta$ and an epistemic state (or belief type) $e \in E$ to any game $G \in \mathcal{G}$.\footnote{Given a game $G$, the set $ \Theta $ is the set of all subjective preferences, or subjective utility functions, on $G$. I.e., $\Theta$ is the set of all functions $\theta: A \rightarrow \mathbb{R}$.} 
Then, given a set of subjective
  preferences $ \Theta $ and
  a set of or belief types $E$, an action selection function is a function
  $\kappa: \Theta \times E \rightarrow A_i $, that associates any subjective representation of the game $ (\theta,e) \in \Theta \times E$ with an action $a^{*} \in A_i $.

More concretely, we focus on two (\textit{subjective}) \emph{preference types} and two \emph{belief
    types}. We view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and
  a belief type as a general disposition to form beliefs about the co-player’s behavior. The
  two possible dispositions that we take account of are precise probabilistic belief and
  imprecise belief, as we will explain in the next section.

Since we are most interested here in studying whether evolutionary pressure could
  support non-veridical subjective preferences, the two preference types that we consider are: the veridical type $\tau^\pi$, who uses the actual payoff function
  $\pi$ of the actual game $G$, and the regret type $\tau^{reg}$, who looks at the regret
  transformation of the actual game $G$.

Finally, the action selection function $\kappa$ that we use in this paper is \textit{maxmin expected
    utility} [GS89], for reasons that will be clear and justified in the following two
  sections.

@@

In evolutionary game theory, for instance, players in a population obviously face uncertainty about the composition of the population that they are part of, and consequently about the co-player that they are randomly paired with at each round and about the co-player's action. We represent this uncertainty in two different ways that correspond to the two approaches discussed above: a Bayesian uniform distribution and an imprecise probability set. This amounts to having two different belief types, $ \overline{\mu}$ and $ \Delta(A_{-i})$, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

@@

In decision theory, the imprecise probability model is in line with most of the representation results of decision making under uncertainty [GilboaSchmeidler89,KMM06], and seems justified by empirical observations too.

@@

The pattern consistently observed in experimets is: $f_{{\color{red}R}} \succ f_{{\color{yellow}Y}}$ and $f_{{\color{red}R}B} \prec f_{{\color{yellow}Y}B}$. This is clearly incompatible with any precise probabilistic belief about the proportion of black and yellow marbles. This kind of behavior has been
famously axiomatized by Gilboa and Schmeidler in a paper from 1989. The choice pattern is normally explained in terms of aversion to uncertainty, and the standard representation is given by means of a set of probability measures together with maxmin rule. We can think of the DM in Ellsberg's example as maximinimizing expected utility over the probability set $ [\underline{P}(R)=\overline{P}(R)=\frac{1}{3},\underline{P}(Y)=0,\underline{P}(B)=0,\overline{P}(Y)=\frac{2}{3},\overline{P}(B)=\frac{2}{3}]  $. This amounts to calculating for any probability measure contained in the probability set the expected utility of any available action, and then to choosing the action that guarantees the highest minimal expected utility. The following observation is then trivial.

@@

Evidence from experimental literature suggests that agents are mainly uncertainty
averse [REF]. In line with empirical data,
we assume that players in the population are uncertainty averse and conform to maxmin expected
utility. The resulting behavior is then produced by maximinimizing the subjective utility
given by the player's preference type over the set of probabilities given by the belief type.

@@

By fixing the action selection component to maximin expected utility, and allowing two belief and
two preference types, we regain all the choice rules considered in sectionNN from a more general and unified standpoint, as it is summed up in
Table~\ref{tab:CMs}. \\
We are now in the position of formally deriving behavior from a pair $(\theta, e) \in \Theta \times E$.

@@

Given a game $G$, a preference type $\tau$ and a belief type $e$, we define $\kappa(\tau(G), e):= \lbrace a_i \in A_i : a_i \in \text{max}_{a'_i \in A_i} \text{ min}_{\mu \in e} \text{ } \tau(G)(a'_i, \mu) \rbrace$.\footnote{As already specified in the previous example, we assume that, whenever the set $\kappa(\tau(G), e)$ is not a singleton, the player $(\tau, e)$ will mix with equal probability among the actions in $\kappa(\tau(G), e)$.}

@@

[[\noindent Definition \ref{def:expected fitness} gives the expected evolutionary fitness of player $(\tau, e)$ against player $(\tau', e')$ in game $G$.
[OF COURSE we will postpone the definition of evolutionary fitness, it is only momentarily here!]
\begin{definition} \label{def:expected fitness}

Given a game $G \in \mathcal{G}$, a player $(\tau, e)$, and a player $(\tau', e')$, we define $\pi^G((\tau, e),(\tau', e')):= \pi^G(\kappa(\tau(G), e),\kappa(\tau'(G), e'))$.

@@

It is then worth stressing here that $\kappa(\tau^{\pi}, \Delta(A_{-i}))$ turns out to be the actual payoff
maximinimizer, $\kappa(\tau^{reg}, \Delta(A_{-i}))$ corresponds to the regret minimizer,
$\kappa(\tau^{\pi}, \overline{\mu})$ is the expected actual payoff maximizer (with flat beliefs), and
$\kappa(\tau^{reg}, \overline{\mu})$ is the expected regret minimizer (with flat beliefs).
Moreover, different subjective preferences and belief types can give rise to identical action predictions,
either in a restricted class of games, or in general. This is also the case here.

@@

\noindent In other words, given the same probabilistic belief, maximization of expected
(objective) payoff and minimization of expected regret are behaviorally
equivalent. This holds for arbitrary games, as well as for solitary decision situations. On the
other hand, the following does not hold in general:

@@

Many other alternatives different from maxmin could produce behavior: choice rules like minmax, maxmax, and minmin would similarly generate behavior in case of a lack of precise probabilistic belief. As it should be clear by now, here we postulate that players use maxmin for two main reasons. First, among other possible choice rules maxmin is not only intuitively appealing but also corroborated from an empirical point of view. Secondly, our main interest is the evolution of different patterns of subjective conceptualizations of games, and we want to show that the veridical representation is not necessarily the fittest and may not be evolutionarily stable. Assuming that all players use a fixed rule (maxmin expected utility) independent of their preference and belief type allows us to factor the choice rule out of the selection process and to focus on subjective representations only.

@@

\todo[inline]{PAOLO rewrites up to here!!! }

@@
